---
title: 'Lab 10: Halloween Mini-Project'
author: "Taylor Darby"
date: "10/29/2021"
output: 
  pdf_document:
    latex_engine: xelatex
---

# Lab 10: Halloween Mini-Project

# Import the data 'candy-data.csv'

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

> **Q1.** How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> **Q2.** How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

## 2. What is your favorite candy

```{r}
candy["Snickers",]$winpercent
```

> **Q3.** What is your favorite candy in the dataset and what is it’s winpercent value?




> **Q4.** What is the winpercent value for “Kit Kat”?




> **Q5.** What is the winpercent value for “Tootsie Roll Snack Bars”?


There is a useful skim() function in the skimr package that can help give you a quick overview of a given dataset. Let’s install this package and try it on our candy data.

```{r}
# install.packages("skimr")
library("skimr")
skim(candy)
```

```{r}
hist(candy$winpercent)
```

> **Q6.** Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?





> **Q7.** What do you think a zero and one represent for the candy$chocolate column?


> **Q8.** Plot a histogram of winpercent values




> **Q9.** Is the distribution of winpercent values symmetrical?



> **Q10.** Is the center of the distribution above or below 50%?



> **Q11.** On average is chocolate candy higher or lower ranked than fruit candy?

### Chocolate is ranked higher

```{r}
# change the chocolate column to a logical (returns true or false)
#  
chocolate <- candy[as.logical(candy$chocolate),]$winpercent
mean(chocolate)
```
```{r}
# same but for fruity
fruity <- candy[as.logical(candy$fruity),]$winpercent
mean(fruity)
```


> **Q12.** Is this difference statistically significant?

### Yes, the p-value = 2.871e-08

```{r}
t.test(chocolate,fruity)
```


## 3. Overall Candy Rankings

Let's make a barplot of the winpercent values for the various candy types

> **Q13.** What are the five least liked candy types in this set?

```{r}

```


1) Chiclets
2) Boston baked beans
3) Nik L Nip

> **Q14.** What are the top 5 all time favorite candy types out of this set?

1) Reese's peanut butter cups
2) Reese's miniatures
3) Twix
4) Kit-Kat
5) Snickers

> **Q15.** Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> **Q16.** This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

We need to improve this to reorder the candy by winpercent values
```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```


Time to add some color:

Create a vector of colors to use to color the plot
```{r}
# Color vector. All black to start
my_cols=rep("black", nrow(candy))

# Now overwrite the black with colors by candy type
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
my_cols
```

Plot using color
```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols)
```

> **Q17.** What is the worst ranked chocolate candy?

Nik L Nip

> **Q18.** What is the best ranked fruity candy?

Reese's Peanut Butter Cups

## 4. Taking a look at pricepercent


```{r}
#install.packages(ggrepel)
## 'ggrepel' helps read labels more easily by repelling them away from each other
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

```{r}
# fix font issues
rownames(candy) <- gsub("Ã","'", rownames(candy))
```


> **Q19.** Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Out of the top 5 candies Reese's Miniatures are the least expensive.


> **Q20.** What are the top 5 most expensive candy types in the dataset and of these which is the least popular?



```{r}
# Correlation analysis
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

```{r}
# PCA Analysis
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```


```{r}
# Now we can plot our main PCA score plot of PC1 vs PC2.
plot(pca$x[,1:2])
```

```{r}
# We can change the plotting character and add some color:

plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

# Same plot as before but colored by candy type and sized by popularity
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
p

# Now we'll label the plot and the data points
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

If you want to see more candy labels you can change the max.overlaps value to allow more overlapping labels or pass the ggplot object p to plotly like so to generate an interactive plot that you can mouse over to see labels:

```{r}
library(plotly)
# Note: hid this plot to more easily knit my pdf for submission
# ggplotly(p)
```


Let’s finish by taking a quick look at PCA our loadings. Do these make sense to you? Notice the opposite effects of chocolate and fruity and the similar effects of chocolate and bar (i.e. we already know they are correlated).

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> **Q24.** What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?




