---
title: 'Lab 17: Vax Mini Project'
author: "Taylor Darby"
date: "11/27/2021"
output: pdf_document
---

# Getting Started

```{r}
# Import vaccination data
vax <- read.csv("zip_data.csv")
head(vax)
head(sort(vax[,1]))
tail(sort(vax[,1]))
```

**Q1.** What column details the total number of people fully vaccinated? 

### Column 9 labeled 'persons_fully_vaccinated'

**Q2.** What column details the Zip code tabulation area? 

### Column 2 labeled 'zip_code_tabulation_area'

**Q3.** What is the earliest date in this dataset? 

### 2021-01-05

**Q4.** What is the latest date in this dataset?

### 2021-11-23

```{r}
skimr::skim(vax)
```

**Q5.** How many numeric columns are in this dataset? 

### There are 9 numeric columns in this dataset.

**Q6.** Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column? 

### There are `r sum(is.na(vax$persons_fully_vaccinated))` missing values in the 'persons_fully_vaccinated' column.

```{r}
sum(is.na(vax$persons_fully_vaccinated))
```

**Q7.** What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 

###

**Q8.** [Optional]: Why might this data be missing?

















