---
title: "Lab 6 Supplement"
author: "Taylor Darby"
date: "10/20/2021"
output: github_document
---

# This is the supplement to Lab 6: R Functions

## Section 1: Improving analysis code by writing functions
> **A.** Improve this regular R code by abstracting the main activities in your own new function. Note, we will go through this example together in the formal lecture. The main steps should entail running through the code to see if it works,... 

```{r}
# (A. Can you improve this analysis code?
# Answer: I had to make some edits in the code to get it to work correctly (i.e. store the 5 appropriate data and values in the global environment). This is not efficient.

df <- data.frame(a=1:10, b=seq(200,400,length=10),c=11:20,d=NA)
df$a <- (df$a - min(df$a)) / (max(df$a) - min(df$a))
df$b <- (df$b - min(df$b)) / (max(df$b) - min(df$b))
df$c <- (df$c - min(df$c)) / (max(df$c) - min(df$c))
df$d <- (df$d - min(df$d)) / (max(df$d) - min(df$d))
df$a
```

> **A. (continued)** ... simplifying to a core
working code snippet, 

```{r}
x <- df$a
x2 <- (x - min(x)) / (max(x) - min(x))
x2
```

> **A. (continued)** ... reducing any calculation duplication, 

```{r}
# Setting 'min(x)' and 'max(x)' as "range" functions will reduce calculation duplication
x <- df$a
rng <- range(x)
# where the function ends up looking like this...
x2 <- (x - rng[1]) / (rng[2] - rng[1])
x2
```


> **A. (continued)** ... and finally transferring your
new streamlined code into a more useful function for you.

```{r}
# This code streamlined the previous code into a more useful function that accounts for 'NA' values.
rescale <- function(x) {
  rng <- range(x, na.rm=TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
# Example showing function is functional
rescale(1:10)

# Now that we know this function works we can save the function either as an R file and source it each time, or we can extract the function ('Code' -> 'Extract Function')
```


> **B.** Next improve the below example code for the analysis of protein drug interactions by abstracting the main activities in your own new function. Then answer questions 1 to 6 below. It is recommended that you start a new Project in RStudio in a new directory and then install the bio3d package noted in the R code below (N.B. you can use the command 'install.packages("bio3d")' or the RStudio interface to do this). 

```{r}
# Install the bio3d package (Note: there are hashtags preceding the code to avoid installing packages multiple times)
#install.packages("bio3d")
```

> **B. (continued)** ... Then run through the code to see if it works,

```{r}
# Overall Goal: Can you improve this analysis code?
library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug
s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s1, chain="A", elety="CA")
s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b
plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```


> **B. (continued)** ... fix any copy/paste errors before simplifying to a core working code snippet,

```{r}
library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug
s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s3, chain="A", elety="CA")
s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b
plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```

> **B. (continued)** ... reducing any calculation duplication, 

> **B. (continued)** ... and finally transferring it into a more useful function for you.




